Estimation of the perceived time of presence of sources in urban acoustic environment using deep learning techniques

### 2019-03-26 paper/paperDraft.tex
relecture rapide

A first corpus is created to validate the use of simulated sound scenes in the development of pleasantness prediction models. It is based on recordings from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations in the 13th district of Paris. A classification of these scenes is proposed in~\cite{gloaguen2017} in terms of ambiance: \textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}. For each of the 19 recordings 45 seconds of audio in a single channel are extracted that represent well the properties of their respective ambiances, and without single events overwhelming their overall perception. The resulting 45~s extracts are replicated to study perceptual differences between recorded and simulated scenes. The recordings are first annotated in terms of the following properties:

@@

Sound scenes are then simulated using simScene~\cite{rossignol2015} with a database of extracts from recordings of isolated sound sources. This database is constructed from part of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for all other sources. The original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are chosen for perceptual comparison, as they explore diverse real-life situations with respect to environment categorization. All 6 recorded and 19 replicated scenes are normalized so that their playback sound level is the same as measured during recording, with a range from 63.9~dB to 79.4~dB.\\

@@

Adjustments are made on the variance of the considered properties to better cover possible scenarios, and the EBR of voice events, mostly read English recordings, are reduced to improve realism. An additional \textit{place} ambiance is added with properties derived empirically from available data. Diverse new scenarios are then generated by sampling the background and event properties distributions. For each scene the time of presence for the traffic, voice and bird sources is estimated using indicators proposed in~\cite{gontier2018}, and 75 scenes are chosen to maximize pairwise distance in the resulting 3-dimensional space. This is shown in Figure~\ref{fig:tvb_pres}. The playback sound level is also conditioned on the ambiance according to typical sound levels in urban context shown in Figure~\ref{fig:amb_levels}, and ranges from 46.6~dB to 77.1~dB.\\

The experiment corpus thus totals 100 scenes: 6 recorded and 19 replicated scenes from the Grafic project, and 75 simulated scenes.

@@

A total of 23 students aged 22 to 23 including 16 males and 7 females at Ecole Centrale de Nantes completed the test. All gave consent prior to the experiment.

@@

\begin{itemize}
\item Z-weighted $L_{eq}$ and A-weighted $LA_{eq}$ equivalent sound levels in dB and dBA respectively.
\item $L_{10}$, $L_{50}$ and $L_{90}$: 10th, 50th and 90th percentiles of the Z-weighted sound level. The $L_{10}$ is associated to events and the $L_{90}$ to background activity, while the $L_{50}$ is a measurement of the overall sound level.
\item $LA_{50}$: 50th percentile of the A-weighted sound level, with similar properties as the $L_{50}$.
\item $L_{50, 1kHz}$: 50th percentile of the Z-weighted sound level for the 1kHz frequency band, also a good descriptor of the overall sound level of the scene.
\item $LA_{10}-LA_{90}$: Emergence indicator included in the pleasantness model in~\cite{ricciardi2014}.
\end{itemize}

@@

The sensors implemented as part of the Cense project provide fast measurements of sound levels for each third-octave band in the $20~Hz - 12.5~kHz$ range in their normal operation~\cite{ardouin2018, gontier2017}. Thus, a prediction model should take as input the same representation. Third-octave spectra are computed for 125~ms frames and 29 frequency bands of the mixed sound scenes. Inputs are then obtained by splitting the resulting spectrograms into frames with a 1~s duration, which is considered a relevant scale for perception. The input frames of dimension 29x8 are then processed independently in this study.\\

Target outputs are associated to each input frame to train the model. Based on previous work the time of presence of traffic, voices and birds can be approximated by the $\hat T_s(\alpha, \beta)$ indicator. Thus, the source presence values before averaging along time can be used as weak labels. Alternatively, it is possible to perform sound level estimation using ground truth labels. Though, this regression task is difficult to learn compared to source detection, and did not result in better time of presence estimations in our experiments.

@@

A deep learning model is implemented and trained for the source presence prediction task, with the architecture shown in Figure~\ref{fig:deep_arch}. The model includes 4 blocks of convolutional layers followed by LeakyReLU activations. The convolutional layers have respectively 128, 64, 32 and 32 output maps, \textit{i.e.} channels, and a common kernel size of 5x5. The output of the last block is flattened then goes through a fully connected layer with output size 3. A final sigmoid activation is used to obtain outputs in the 0-1 range. The resulting 3 values correspond to the presence of traffic, voices and birds respectively, and are compared during training to presence labels given by $\hat T_s(\alpha, \beta)$ using a binary cross-entropy cost function:
\begin{equation}
L(y, \hat y) = -\sum_s y_s log\left(\hat y_s\right) + (1-y_s) log\left(1-\hat y_s\right)
\end{equation}
where $s$ is the source, $y_s$ and $\hat y_s$ are the target and predicted presence for source $s$ in the 0-1 range. This loss function is minimized using the Adam algorithm~\cite{kingma2015} on batches of 1~s examples.\\

@@

To validate the use of simulated sound scenes with new scenarios as well as reduced source complexity in the development of prediction models, the perceptual space generated by the experiment's five general scales is studied. It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses at the scene level. Figure~\ref{fig:pspace} compares the results for scenes based on recordings (left, n=25) and new scenarios (right, n=75). The generated spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two corpora. For both sets the variance explained by the first two components is similar (resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%). Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson2010, cain2013}. Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities. Additionally, assessments for each scene are projected onto the PCA space and represented in Figure~\ref{fig:pspace} as dots. The space covered by scenes based on new scenarios covers that of the studied real-life environments, further demonstrating the efficiency of the scene generation procedure in terms of diversity.\\

Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace}a by arrows, and the standard deviation of assessments is represented by ellipses for one scene. The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes, not shown here for clarity. This illustrates the results in Table~\ref{tab:ogrep} where perceptual assessments were not found to differ significantly overall.\\

% FIGURE
\begin{figure}[h]
    \centering
     \begin{subfigure}[t]{0.55\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/pca_p1.eps}
    \end{subfigure}%
    \begin{subfigure}[t]{0.5\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/pca_sim.eps}
    \end{subfigure}
    \caption{Principal components analysis (first two components) of the four general questions at the assessment level (n=252). Each dot represent the projection of one scene on the resulting space. The components space is distorted compared to the litterature.}\label{fig:pspace}
\end{figure}

@@

Next, the capacity of these predictions to result in relevant pleasantness estimates is studied. For this analysis only the perceptual experiment is used as no assessments are available on the evaluation dataset. To evaluate the model's robustness to the increased polyphony and source complexity of scenes in real-life scenarios, the corpus is further split in two parts:
\begin{enumerate}
\item The 25 recorded and replicated scenes, which contain additional sources not present in the development or evaluation datasets simulated for the optimization of deep learning models,
\item The 75 simulated scenes that are obtained from the same simulation process as both the development and evaluation dataset.
\end{enumerate}
Pleasantness predictions are obtained by substituting time of presence estimates computed from the presence detection architecture's outputs to the $\hat P_{\varphi 1}$ model presented in Section~\ref{sec:base_phys}. Thus, only the outputs corresponding to the presence of birds are used. The results are compared to estimates using ground truth $\hat T_s(\alpha, \beta)$ labels in Table~\ref{tab:pppred}. First, the metrics computed on the whole test corpus do not differ significantly for ground truth or predicted labels, with an overall RMSE of about 0.84 on a 10-point scale and 82\% of explained variance in the data. This performance of the detection model is expected given the low errors on time of presence estimates. Labels from the detection model result in higher errors on the first sub-corpus containing recorded and replicated scenes, with sources not seen during training. In both cases the error is still lower on the first sub-corpus, indicating that the detection model generalizes well to new content and sources.\\

@@

Pleasantness prediction using the detection model is as performant as the best baseline model from Section~\ref{sec:base_phys}, and outperforms previous models relying on physical indicators directly computed on the audio signal. Though, errors are significantly higher than for perceptual models in Section~\ref{sec:base_perc}.\\

@@

The perceptual models of pleasantness models are comparable to those proposed in~\cite{ricciardi2014} and~\cite{aumond2017}, of respective expressions (translated from 1-11 scales to 0-10):
\begin{align}
\hat P_{p3} &= 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} + 0.15~T_{B, p}\\
\hat P_{p4} &= 8.70 - 0.47~OL - 0.21~T_{T, p} + 0.12~T_{V, p} + 0.09~T_{B, p}
\end{align}
Although the aim of this work is not to propose a new perceptual model of pleasantness, a few hypotheses can be made from this result. First, the use of a simulated corpus with low source complexity does not alter the fundamental relationships between perceptual quantities, nor does the in vivo setting of the experiment. However, the latter may have an impact on the contribution of the overall loudness to pleasantness: the regression coefficient is $-0.67$, which is significantly higher than for in situ models. This may be due to scenes at the same sound level appearing louder when played through headphones in a laboratory setting.\\

The physical model on this corpus includes only the contribution of bird sources. This is attributed to the overall low diversity of the studied corpus:
\begin{itemize}
\item Only traffic, voice and bird sources are present in simulated sound scenes, as a result global sound level measurements tend to be correlated with the presence of traffic in the simulation process. Quieter environments such as parks and quiet streets are less likely to contain continuous traffic, while high sound levels in busy streets are always attributed to traffic. Including other sources such as construction noises and other transport-related contributions in such environments would lower the observed correlation.
\item The diversity of available isolated source extracts for scene simulation is low. Particularly, voice events are exclusively recordings of read english with low variations in speaker and consistently neutral expressiveness. These properties result in no significant contribution of voice events to pleasantness for this corpus.
\end{itemize}
Thus, in real life scenarios where detection models are applied to sensor networks, indicators related to traffic and voice sources will become relevant for the prediction of pleasantness. The trained deep learning architecture already displays few detection errors for these sources, and can be extended to any additional source provided enough data is available.\\

Results show that pleasantness estimates using the developped source detection architecture are similar to those using ground truth $\hat T_s(\alpha, \beta)$ labels on the simulated corpus. This indicates that detection errors found in Table~\ref{tab:perf_cmp} do not impact pleasantness prediction on average. Because the labels are extracted from a masking model approximating the perceptual time of presence with its own errors, they can be considered "weak" labels. Thus, part of the errors committed by the deep learning model are in fact correct detections compared to wrong labels.\\

The detection-based pleasantness predictions are worse on the corpus containing recorded and replicated scenes, which feature higher polyphonies and a richer source taxonomy. This is expected as the architecture was never trained on complex scenes. Though, interestingly pleasantness estimates are still better on average than on the corpus of simulated scenes. The detection model is thus quite robust to scenes with additional sources as perturbators. A full study with a larger evaluation corpus would be needed to further validate the model's robustness to diverse real-life scenarios.

>>>>

A first corpus is created to validate the use of simulated sound scenes in the development of pleasantness prediction models. It is based on recordings from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations in the 13th district of Paris. A classification of these scenes is proposed in~\cite{gloaguen2017} in terms of ambiance: \textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}. For each of the 19 recordings 45 seconds of audio in a single channel are extracted that represent well the properties of their respective ambiances, without single events overwhelming their overall perception. The resulting 45~s extracts are replicated to study perceptual differences between recorded and simulated scenes. The recordings are first annotated in terms of the following properties:

@@

Sound scenes are then simulated using simScene~\cite{rossignol2015} with a database of extracts from recordings of isolated sound sources. This database is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources. The original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are chosen for perceptual comparison, as they explore diverse real-life situations with respect to environment categorization. All 6 recorded and 19 replicated scenes are normalized so that their playback sound level is the same as measured during recording, with a range from 63.9~dB to 79.4~dB.\\

@@

Adjustments are made on the variance of the considered properties to better cover possible scenarios. The EBR of voice events, mostly read English recordings, are reduced to improve realism. An additional \textit{place} ambiance is added with properties derived empirically from available data. Diverse new scenarios are then generated by sampling the background and event properties distributions. For each scene the time of presence for the traffic, voice and bird sources is estimated using indicators proposed in~\cite{gontier2018}, and 75 scenes are chosen to maximize pairwise distance in the resulting 3-dimensional space. This is shown in Figure~\ref{fig:tvb_pres}. The playback sound level is also conditioned on the ambiance according to typical sound levels in urban context shown in Figure~\ref{fig:amb_levels}, and ranges from 46.6~dB to 77.1~dB.\\

The experiment corpus thus totals 100 scenes: 6 recorded and 19 replicated scenes from the Grafic project \ml{a citer si pas fait avant}, and 75 simulated scenes.

@@

A total of 23 students aged 22 to 23 including 16 males and 7 females at Ecole Centrale de Nantes completed the test. All gave consent prior to the experiment and reported normal hearing.

@@

\begin{itemize}
\item Z-weighted $L_{eq}$ and A-weighted $LA_{eq}$ equivalent sound levels in dB and dBA respectively.
\item $L_{10}$, $L_{50}$ and $L_{90}$: 10th, 50th and 90th percentiles of the Z-weighted sound level. The $L_{10}$ is associated to events and the $L_{90}$ to background activity, while the $L_{50}$ is a measurement of the overall sound level.
\item $LA_{50}$: 50th percentile of the A-weighted sound level, with similar properties as the $L_{50}$.
\item $L_{50, 1kHz}$: 50th percentile of the Z-weighted sound level for the 1kHz frequency band, also a good descriptor of the overall sound level of the scene.
\item $LA_{10}-LA_{90}$: Emergence indicator included in the pleasantness model presented in~\cite{ricciardi2014}.
\end{itemize}

@@

\ml{ Je prendrais un point de vue plus general, tout le monde fait alors on utilise ca comme ca ca montre la capacité de generalisation de l'approche // The sensors implemented as part of the Cense project provide fast measurements of sound levels for each third-octave band in the $20~Hz - 12.5~kHz$ range in their normal operation~\cite{ardouin2018, gontier2017}. Thus, a prediction model should take as input the same representation.} Third-octave spectra are computed for 125~ms frames and 29 frequency bands of the mixed sound scenes. Inputs are then obtained by splitting the resulting spectrograms into frames with a 1~s duration, which is considered a relevant scale for perception. The input frames of dimension 29x8 are then processed independently in this study.\\

Target outputs are associated to each input frame to train the model. Based on previous work the time of presence of traffic, voices and birds can be approximated by the $\hat T_s(\alpha, \beta)$ indicator. Thus, the source presence values before averaging along time can be used as weak labels. Alternatively, it is possible to perform sound level estimation using ground truth labels. Though, this regression task is difficult to learn compared to source detection, and did not result in better time of presence estimations in our experiments. \ml{les resultats seront discutés ensuite ou non ? Sinon le dire ici}

@@

A deep learning model is implemented and trained for the source presence prediction task, with the architecture shown in Figure~\ref{fig:deep_arch}. The model includes 4 blocks of convolutional layers followed by LeakyReLU \ml{un peu technique, utiliser les terms complets et citer} activations. The convolutional layers have respectively 128, 64, 32 and 32 output maps, \textit{i.e.} channels, and a common kernel size of 5x5. The output of the last block is flattened then goes through a fully connected layer with output size 3. A final sigmoid activation is used to obtain outputs in the 0-1 range. The resulting 3 values correspond to the presence of traffic, voices and birds respectively, and are compared during training to presence labels given by $\hat T_s(\alpha, \beta)$ using a binary cross-entropy cost function:
\begin{equation}
L(y, \hat y) = -\sum_s y_s log\left(\hat y_s\right) + (1-y_s) log\left(1-\hat y_s\right)
\end{equation}
where $s$ is the source, $y_s$ and $\hat y_s$ are the target and predicted presence for source $s$ in the 0-1 range. This loss function is minimized using the Adam algorithm~\cite{kingma2015} on batches of 1~s examples.\\

@@

To validate the use of simulated sound scenes with new scenarios as well as reduced source complexity in the development of prediction models, the perceptual space generated by the experiment's five general scales is studied. It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses at the scene level. Figure~\ref{fig:pspace} compares the results for scenes based on recordings (left, n=25) and new scenarios (right, n=75). The resulting spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two corpora. For both sets the variance explained by the first two components is similar (resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%). Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson2010, cain2013}. Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities. Additionally, assessments for each scene are projected onto the PCA space and represented in Figure~\ref{fig:pspace} as dots. The space covered by scenes based on new scenarios covers that of the studied real-life environments, further demonstrating the effectiveness of the scene generation procedure in terms of diversity.\\

Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace} using arrows. The standard deviation of assessments is represented by ellipses for one scene. The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes, not shown here for clarity. This illustrates the results in Table~\ref{tab:ogrep} where perceptual assessments were not found to differ significantly overall.\\

% FIGURE
\begin{figure}[h]
    \centering
     \begin{subfigure}[t]{0.55\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/pca_p1.eps}
    \end{subfigure}%
    \begin{subfigure}[t]{0.5\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/pca_sim.eps}
    \end{subfigure}
    \caption{Principal components analysis (first two components) of the four general questions at the assessment level (n=252). Each dot represent the projection of one scene on the resulting space. The components space is distorted compared to the literature.}\label{fig:pspace}
\end{figure}

@@

Next, the capacity of these predictions to give relevant pleasantness estimates is studied. For this analysis, only the perceptual experiment is used as no assessments are available on the evaluation dataset. To evaluate the model's robustness to the increased polyphony and source complexity of scenes in real-life scenarios, the corpus is further split in two parts:
\begin{enumerate}
\item The 25 recorded and replicated scenes, which contain additional sources not present in the development or evaluation datasets simulated for the optimization of deep learning models,
\item The 75 simulated scenes that are obtained from the same simulation process as both the development and evaluation dataset.
\end{enumerate}
Pleasantness predictions are obtained by substituting time of presence estimates computed from the presence detection architecture's outputs to the $\hat P_{\varphi 1}$ model presented in Section~\ref{sec:base_phys}. Thus, only the outputs corresponding to the presence of birds are used. The results are compared to estimates using ground truth $\hat T_s(\alpha, \beta)$ labels in Table~\ref{tab:pppred}. \ml{Je ne comprends pas : First, the metrics computed on the whole test corpus do not differ significantly for ground truth or predicted labels, with an overall RMSE of about 0.84 on a 10-point scale and 82\% of explained variance in the data.} This performance of the detection model is expected given the low errors on time of presence estimates. Labels from the detection model result in higher errors on the first sub-corpus containing recorded and replicated scenes, with sources not seen during training. In both cases the error is still lower on the first sub-corpus, indicating that the detection model generalizes well to new content and sources.\\

@@

Pleasantness prediction using the detection model is as effective as the best baseline model from Section~\ref{sec:base_phys}, and outperforms previous models relying on physical indicators directly computed on the audio signal. Though, errors are significantly higher than the one of perceptual models in Section~\ref{sec:base_perc}.\\

@@

The perceptual models of pleasantness \ml{models} are comparable to those proposed in~\cite{ricciardi2014} and~\cite{aumond2017}, of respective expressions (translated from 1-11 scales to 0-10):
\begin{align}
\hat P_{p3} &= 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} + 0.15~T_{B, p}\\
\hat P_{p4} &= 8.70 - 0.47~OL - 0.21~T_{T, p} + 0.12~T_{V, p} + 0.09~T_{B, p}
\end{align}
\ml{Enlever ou reformuler: Although the aim of this work is not to propose a new perceptual model of pleasantness, a few hypotheses can be made from this result.} First, the use of a simulated corpus with low source complexity does not alter the fundamental relationships between perceptual quantities, nor does the in vivo setting of the experiment. However, the latter may have an impact on the contribution of the overall loudness to pleasantness: the regression coefficient is $-0.67$, which is significantly higher than for in situ models. This may be due to scenes at the same sound level appearing louder when played through headphones in a laboratory setting.\\

The physical model on this corpus includes only the contribution of bird sources. This result is attributed to the overall low diversity of the studied corpus:
\begin{itemize}
\item Only traffic, voice and bird sources are present in simulated sound scenes, as a result global sound level measurements tend to be correlated with the presence of traffic in the simulation process. Quieter environments such as parks and quiet streets are less likely to have continuous traffic, while high sound levels in busy streets are always due to busy traffic. It is expected that including other sources such as construction noises and other transport-related contributions in such environments would lower the observed correlation.
\item The diversity of available isolated source extracts for scene simulation is rather low. Particularly, voice events are recordings of read english with low variations in speaker and consistently neutral expressiveness. These properties result in no significant contribution of voice events to pleasantness for this corpus.
\end{itemize}
Thus, in real life scenarios where detection models are applied to sensor networks, indicators related to traffic and voice sources will become relevant for the prediction of pleasantness. The trained deep learning architecture already displays few detection errors for these sources, and can be extended to any additional source provided enough data is available.\\

Pleasantness estimates using the proposed source detection architecture are similar to those using ground truth $\hat T_s(\alpha, \beta)$ labels on the simulated corpus. This indicates that detection errors found in Table~\ref{tab:perf_cmp} do not impact pleasantness prediction on average. Since the labels are extracted from a masking model approximating the perceptual time of presence with its own errors, they can be considered as "weak" labels. Thus, part of the errors committed by the deep learning model are in fact correct detections compared to wrong labels.\ml{faut faire attention au wording, je n'utiliserai pas bon et mauvais, mais plus le fait que les labels sont plus ou moins haut niveau ?}\\

The detection-based pleasantness predictions are of lower quality on the corpus containing recorded and replicated scenes, which feature higher polyphonies and a richer source taxonomy. This is expected as the architecture is not trained on complex scenes. Though, interestingly pleasantness estimates are still better on average than on the corpus of simulated scenes. The detection model is thus quite robust to scenes with additional sources as perturbators. A full study with a larger evaluation corpus would be needed to further validate the model's robustness to diverse real-life scenarios.
